---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

# Predictive probability of success (PPoS) - application to time-to-event data in the presence of competing risks

Chiara Micoli, Andrea Discacciati

Karolinska Institutet, Sweden

Published: 2024-08-12

Last updated: `r Sys.Date()`

------------------------------------------------------------------------

Reproducible example using simulated data for the the paper "Simulation-based Bayesian predictive probability of success for interim monitoring of clinical trials with competing event data: two case studies".

------------------------------------------------------------------------

Functions can be `source`’d directly from R:

```{r}
source("https://raw.githubusercontent.com/cmicoli/PPOS-CompetingRisks/main/00Functions_weibull_PPOS.R")
source("https://raw.githubusercontent.com/anddis/brms-weibullPH/main/weibullPH_funs.R")
```

```{r, message=FALSE, warning=FALSE}
RNGkind("L'Ecuyer-CMRG")
set.seed(123)

library(brms)
library(survival)
library(flexsurv)
library(tidybayes)
library(tidyverse)
library(parallel)
```

```{r}
theme_set(theme_bw())
```

### Simulate right-censored survival data.

Simulating data until interim analysis, corresponding to an interim data collection.

```{r}
dat <- local({
  set.seed(1901)
  N <- 150
  TR <- rbinom(N, 1, 0.45)
  c <- rbinom(N, 1, 0.2)
  z <- rnorm(N)
  y <- rweibullPH(N, 
                  shape =  1.5, # gamma
                  scale = exp(-6 + log(1.4)*TR + log(0.8)*z)) # mu
  cens <- runif(N, 0, 60)
  time <- pmin(y, cens)
  status <- as.numeric(y <= cens)
  
  event <- ifelse(status == 1, rbinom(N, 1, 0.3)+1, 0) #competing events 1-2
  
  data.frame(
    ID = 1:N,
    TIME = time,
    status = status,
    censored = 1 - status,
    EVENT = event,
    TREATMENT =  factor(TR,
                        levels = c(0, 1),   
                        labels = c("CNTR", "NEW_TR")), #factor(TR),
    c = factor(c),
    z = z
  )
})

head(dat)
```

Presenting event summary for controls and new treatment group: event 1 and event 2 are the two competing events, 0 for censored data.

```{r}
table(dat$TREATMENT,  useNA = "ifany")
table(dat$TREATMENT, dat$EVENT, useNA = "ifany")
summary(dat$TIME)
```

Preparing data for fitting brms models.

```{r}
dat$event.all <- ifelse(dat$EVENT %in% c(1,2), 1, 0)
dat$event1 <- dat$event.all * (dat$EVENT == 1)
dat$event2 <- dat$event.all * (dat$EVENT == 2)
dat$censored1 <- 1 - dat$event1
dat$censored2 <- 1 - dat$event2
```

## Definition of success

Time to recovery is analyzed in the trial by using a Weibull distribution with prespecified weakly informative priors. To do so event 1 ($i=1$) and event 2 ($i=2$) are considered as competing events, and two separate Bayesian proportional-hazard Weibull models will be used, one for each competing outcome $i$. For each model, we will let the logarithm of the Weibull distribution’s scale parameter u to depend on treatment arm (X; new treatment or control arm) and on baseline value of C, so that $log(u_i) = \alpha_i + \beta_i X + \gamma_i C$. The distribution’s shape parameter ($v^i$), which is constrained to be positive, will not depend on any covariates. The following priors are used in the primary analysis:

$\alpha_i \sim Normal (0, 20)$

$\beta_i \sim Normal (0, \sqrt{0.5})$

$\gamma_i \sim Normal (0, \sqrt{0.5})$

$v_i \sim Exponential(1)$

Success is here defined as achieving the probability threshold for the new treatment to graduate to the next phase of drug development. The rule for graduation is formalized as $P(csHR_{event1} > 1 | Data) \geq 0.975$.

## Analysis at interm

Start by fitting Fit Weibull PH model with brms. Model to fit: we consider treatment arm and one binary covariate (c) in the models.

```{r}
## models 
bbff1 <- bf(TIME | cens(censored1) ~   0 + Intercept + TREATMENT + c,
            family = weibullPH)
bbff2 <- bf(TIME | cens(censored2) ~   0 + Intercept + TREATMENT + c,
            family = weibullPH)
```

```{r}
## output is median HR, and 95% qCrI for treatment effect, 
## and probability of recovery graduating : p(HR >1 | Dati) = # draws with HR >1 / 4000
fit.main.event1 <- brm(bbff1,
                       data = dat,     
                       chains = 4,
                       iter = 1500,
                       warmup = 500,
                       seed = 12345,
                       cores = 1,
                       stanvars = stanvars_weibullPH,
                       prior = c(prior(normal(0, 20), class = "b", coef = "Intercept"),
                                 prior(normal(0, sqrt(.5)), class = b), 
                                 prior(exponential(1), class = gamma)),
                       backend = "cmdstanr", ## rstan
                       refresh = 0)
fit.main.event2 <- brm(bbff2,
                       data = dat,     
                       chains = 4,
                       iter = 1500,
                       warmup = 500,
                       seed = 12345,
                       cores = 1,
                       stanvars = stanvars_weibullPH,
                       prior = c(prior(normal(0, 20), class = "b", coef = "Intercept"),
                                 prior(normal(0, sqrt(.5)), class = b), 
                                 prior(exponential(1), class = gamma)),
                       backend = "cmdstanr", ## rstan
                       refresh = 0) 
```

```{r}
#fit.main.event1
pos.draws <- fit.main.event1 %>% 
  spread_draws(b_Intercept, b_TREATMENTNEW_TR , b_c1   , gamma)
# print(dim(pos.draws))
# print(colSums(is.na(pos.draws)))
HR.CrI <- quantile(exp(pos.draws$b_TREATMENTNEW_TR),
                   probs =c(0.5, 0.025, 0.975),  na.rm = FALSE) %>% 
  as.data.frame() %>% t() 
HR.CrI.pr <- HR.CrI %>% 
  cbind(count.HRvs1  = sum(exp(pos.draws$b_TREATMENTNEW_TR)>1)) %>% 
  as.data.frame() %>%
  mutate(pr = count.HRvs1 /dim(pos.draws)[1], ## probability of graduating in this dataset
         success = ifelse(pr > 0.975, 1, 0 )) ## definition of success with this dataset
HR.CrI.pr
```

# Predictive probability of success analysis

### 1. Splitting observed data

Administrative censoring after 60 days.

```{r}
# splitting between completely observed and censored data
dat1 <- dat %>% filter(event.all == 1 | TIME  == 60) %>% 
  select(ID, TREATMENT, c, TIME, EVENT)  ## 101 patients
dat0_cens <- dat %>% filter(event.all == 0 & TIME < 60) ## 32 patients
```

### 2. Defining number simulations K

```{r}
nsim <- 100  ## K
# cores to use in the fitting if parallelized
cores <- 4
```

### 3. Predicting baseline allocation of new patients

Maximum enrolment for the new treatment arm is 125.

```{r}
extra.new.tr <- 125 - sum(dat$TREATMENT == "NEW_TR")
## new treatment has to be present 60 times, X is the number of controls needed to enroll 60 new_tr
D_new <- lapply (1:nsim, function(x) {
  n.control <- rnbinom(1, extra.new.tr, 0.45)
  d <- data.frame(
    ID = paste0("NEW", seq_len(n.control+ extra.new.tr)),
    TREATMENT = factor(rep(c("CNTR", "NEW_TR"), times = c(n.control, extra.new.tr))),
    c = factor(rbinom(n.control+extra.new.tr, 1, 
                      rbeta(1, table(dat$c)["1"]+1, sum(table(dat$c))-table(dat$c)["1"]+1)),
               levels = c(0, 1))
  )
})

# dat0 contains data that need prediction, either update of follow up for dat0_cens, or for D_new
dat0 <- lapply (D_new, function(x) {
  dat0 <- dat0_cens %>% 
    select(ID, TREATMENT, c, TIME, EVENT) %>%
    rbind (x %>% 
             mutate(TIME = 0, EVENT = 0))
})

```

Splitting into control and new treatment dataset, for future steps where we will fit separate models

```{r}
# new treatment datasets
dat_new.tr1 <- dat %>% filter(TREATMENT == "NEW_TR") %>% 
  filter(event.all == 1 | TIME  == 60) %>% 
  select(ID, TREATMENT, c, TIME, EVENT)  ## 45 patients
dat_new.tr0 <- lapply(dat0, \(x) x %>% filter(TREATMENT == "NEW_TR") )
#function(x) {  dat_new.tr0 <-  x %>% filter(TREATMENT == "NEW_TR")})
## 89 patients to predict: 32 already enrolled, 60 new

# control datasets
dat_control1 <- dat %>% filter(TREATMENT == "CNTR") %>% 
  filter(event.all == 1 | TIME  == 60) %>% 
  select(ID, TREATMENT, c, TIME, EVENT)  ## 55 patients
dat_control0 <- lapply(dat0, \(x) x %>% filter(TREATMENT == "CNTR") )
## XX patients to predict: 22 already enrolled, new patients in the trial (new accruals - from negative binomial)
```

## Main analysis (A)

Modeling stratifying over treatment arm, using C as the only covariate in the model.

### 4.A - Modeling step

```{r}
## models 
bbff1.modelingA <- bf(TIME | cens(censored1) ~   0 + Intercept  + c,
                      family = weibullPH)
bbff2.modelingA <- bf(TIME | cens(censored2) ~   0 + Intercept  + c,
                      family = weibullPH)
```

```{r}
# model fit for new treatment arm
fit_new.tr <- fit.noPH(dataset = dat %>% filter(TREATMENT == "NEW_TR"), 
                       brms_function1 = bbff1.modelingA,
                       brms_function2 = bbff2.modelingA   )

# model fit for control arm 
fit_control <- fit.noPH(dataset = dat %>% filter(TREATMENT == "CNTR"), 
                        brms_function1 = bbff1.modelingA,
                        brms_function2 = bbff2.modelingA   )
```

## 5.A - Predictive step

```{r}
## new predicted dataset 
predicted_new.tr <- predicted_dataset_noPH(data_to_predict = dat_new.tr0, 
                                           data_observed = dat_new.tr1, 
                                           n.sim = nsim, 
                                           model1 = fit_new.tr$Fit1, 
                                           model2 = fit_new.tr$Fit2, 
                                           cores = cores)

predicted_control <- predicted_dataset_noPH(data_to_predict = dat_control0, 
                                            data_observed = dat_control1, 
                                            n.sim = nsim, 
                                            model1 = fit_control$Fit1, 
                                            model2 = fit_control$Fit2, 
                                            cores = cores)

predicted <- lapply (1:nsim , \(x) rbind(predicted_control[[x]] , 
                                         predicted_new.tr[[x]]))
```

## 6.A - Analytic step

```{r, message =FALSE}
## analysis of new predicted dataset 
fit.null_analysis <- compile.analysis.model(predicted[[1]])

predicted_an <- mclapply(predicted, 
                         \(x)  main.analysis.update(dataset = x, fit.null = fit.null_analysis),
                         mc.cores = cores)
```

```{r}
analysis.HR <- as.data.frame(do.call(rbind, predicted_an %>% map(2)))
analysis.HR$success <- factor(analysis.HR$success, levels = 0:1)
head(analysis.HR)
```

```{r}
PPOS <- (table(analysis.HR$success, useNA = "ifany") %>% prop.table())["1"]
PPOS
```

## Sensitivity analysis (B)

Modeling using PH assumption on the treatment effect and adding prior belief on treatment effect, using C as extra covariate in the model.

Choosing $\mu$ (mu.fit) and $\sigma^2$ (var.fit) for prior on the treatment effect

```{r}
mu.fit <- 2
var.fit <- 0.1 
```

### 4.B - Modeling step

Prior on log-treatment effect (log(csHR)) centered on $log(2)$
```{r}
## models 
bbff1.modelingB <- bf(TIME | cens(censored1) ~   0 + Intercept + TREATMENT + c,
                      family = weibullPH)
bbff2.modelingB <- bf(TIME | cens(censored2) ~   0 + Intercept + TREATMENT + c,
                      family = weibullPH)
```


```{r}
# model fit 
fit_HR2 <- fit.PH.HR(dataset = dat, 
                     brms_function1 = bbff1.modelingB,
                     brms_function2 = bbff2.modelingB,
                     logHR.mean = log(mu.fit), 
                     logHR.sd = sqrt(var.fit))
```

### 5.B - Predictive step

```{r}
pred.datasets_HR2 <- predicted_dataset_PH.HR(data_to_predict = dat0, 
                                             data_observed = dat1, 
                                             model1 = fit_HR2$Fit1, 
                                             model2 = fit_HR2$Fit2, 
                                             n.sim = nsim , cores)
```

### 6.B - Analytic step

```{r, message =FALSE}
fit.null_analysis <- compile.analysis.model(pred.datasets_HR2[[1]])

analysis.datasets_HR2 <- mclapply(pred.datasets_HR2, 
                                  \(x) main.analysis.update(x, fit.null_analysis), mc.cores = cores)
```

```{r}
analysis.HR2 <- as.data.frame(do.call(rbind, analysis.datasets_HR2 %>% map(2)))
analysis.HR2$success <- factor(analysis.HR2$success, levels = 0:1)
```

```{r}
table(analysis.HR2$success, useNA = "ifany")

PPOS <- (table(analysis.HR2$success, useNA = "ifany") %>% prop.table())["1"]
PPOS
```

# Session info

```{r}
sessionInfo()
```
